---
title: "Sunrise and sunset"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Whereas the durations of day and night as a function of time follow a sinusoid
function, this is not the case for the timing of the sunrise and sunset because
of the facts (i) that the orbit of the earth around the sun is an ellipse and
not a circle and (ii) the rotation axis of the earth is tilted on the orbital
plan. Here we illustrate all this in a given location.

## Packages

```{r message = FALSE}
library(hms)
library(dplyr)
library(suncalc)
library(lubridate)
library(purrr)
library(magrittr)
```

## Parameters:

```{r}
saigon <- list(lat =  10.802506629026347,
               lon = 106.73305962858632,
               tz  = "Asia/Saigon")

oxford <- list(lat =  51.75187551287947,
               lon =  -1.2561729436820785,
               tz  = "Europe/London")
```

## Functions

```{r}
plot2 <- function(...) plot(..., type = "l", lwd = 2)
lines2 <- function(...) lines(..., lwd = 2)
```


## Computing the sunrise and sunset times:

A function that computes the times of sunrise and sunset in a given location and
between 2 dates:

```{r}
sunriseset <- function(where, from, to, by = 1) {
  seq.Date(from, to, by) |> 
    getSunlightTimes(lat = where$lat, lon = where$lon) |>
    select(date, sunrise, sunset) |> 
    mutate(across(-date, ~ as_hms(with_tz(.x, where$tz)))) |> 
    mutate(day_duration = sunset - sunrise) |> 
    as_tibble()
}
```

Let's try it:

```{r}
from <- ymd(20230101)
to <- ymd(20241231)
saigon_val <- sunriseset(saigon, from, to)
oxford_val <- sunriseset(oxford, from, to)
```

## Looking at the duration of the day

```{r}
with(oxford_val, plot2(date, day_duration / 60 / 60, col = 2, xlab = "",
                       ylab = "day duration (hours)"))

with(saigon_val, lines2(date, day_duration / 60 / 60, col = 3))

abline(v = ymd(20240101))
abline(h = 12)
```


```{r}
f <- function(x) map_dfr(c(min, max),
                         ~ first(filter(x, day_duration == .x(day_duration))))

g <- function(x) x |>
  pull(day_duration) |>
  mean() |> 
  divide_by (60 * 60) |> 
  as.vector()

(sgn <- f(saigon_val))
(oxf <- f(oxford_val))

g(sgn)
g(oxf)
```

When is it that the durations of the day in Oxford and Saigon are exactly the
same?

```{r}
left_join(saigon_val, oxford_val, "date") |> 
  mutate(diff = abs(day_duration.x - day_duration.y)) |> 
  arrange(diff)
```

## Timings of sunrise and sunset

```{r}
saigon_val |>
  arrange(sunrise) |> 
  select(date, sunrise) |> 
  first()

saigon_val |>
  arrange(desc(sunrise)) |> 
  select(date, sunrise) |> 
  first()

saigon_val |>
  arrange(sunset) |> 
  select(date, sunset) |> 
  first()

saigon_val |>
  arrange(desc(sunset)) |> 
  select(date, sunset) |> 
  first()

oxford_val |>
  arrange(sunrise) |> 
  select(date, sunrise) |> 
  first()

oxford_val |>
  arrange(desc(sunrise)) |> 
  select(date, sunrise) |> 
  first()

oxford_val |>
  arrange(sunset) |> 
  select(date, sunset) |> 
  first()

oxford_val |>
  arrange(desc(sunset)) |> 
  select(date, sunset) |> 
  first()
```

```{r}
with(saigon_val, {
  plot2(date, sunrise / 60 / 60, col = 2, ylim = c(5, 19),
        ylab = "time of daylight")
  lines(date, sunset / 60 / 60, col = 4, lwd = 2)
  lines(date, sunset / 60 / 60 - 1, col = 4, lwd = 2, lty = 2)
  lines(date, sunrise / 60 / 60 + 1, col = 2, lwd = 2, lty = 2)
})

abline(v = ymd(20240101))

abline(h = c(7, 8, 17, 18))
```

An alternative way of representing:

```{r}
with(saigon_val, {
  plot2(sunrise / 60 / 60, date, col = 2, xlim = c(5, 19),
        xlab = "time of daylight")
  lines(sunset / 60 / 60, date, col = 4, lwd = 2)
  lines(sunset / 60 / 60 - 1, date, col = 4, lwd = 2, lty = 2)
  lines(sunrise / 60 / 60 + 1, date, col = 2, lwd = 2, lty = 2)
})

abline(h = ymd(20240101))

abline(v = c(6, 7, 17, 18))
```





